---
title: "Slide 01: Data"
author: "Richard Dien Giau Bui"
format: revealjs
editor: visual
---

## Load packages

-   First, before a code session, think which packages you need to use today

    -   Like before cooking, which knifes we should use

-   Load it with `library(name_of_package)`:

```{r}
#| echo: true
#| warning: false
library(tidyverse)
library(skimr)
```

## Code block

-   Insert a new code block with **CTRL+ALT+I** then run with the green **RUN** button

```{r}
#| echo: true
60 + 8
```

## Syntax of R language

-   Like any language, R also has their own syntax:

```{r}
#| eval: false
#| echo: true
OUTPUT = FUNCTION(INPUT, OTHER_OPTIONS)
```

-   For example: we take function `sqrt` for number 4 (as input) and the results will be saved to new object `y` . How much is `y` ?

```{r}
y = sqrt(4)
```

## Comments in R

-   Some notes to our codes

-   Not run and just give information

```{r}
#| echo: true
# two ways to take square of a number in R: 
3*3 
3^2
```

## Four main data types in R

-   `numeric`: for example, counting, km, scores, money, ...

-   `character`: for example, "Wednesday", "Today I failed the coding class", ...

-   `factor`: e.g., a column with only 4 possible categories: A, B, C, D

-   `date`: `2024-02-28`

## Type 1. `numeric`

-   Create a numeric variable

```{r}
#| echo: true
x = c(1:10) # other way: x <- c(1:10)
typeof(x)
x = c(0.1, 0.4, 0.5)
typeof(x)
```

-   Some math manipulation with numeric

```{r}
#| echo: true
length(x)
mean(x)
median(x)
min(x)
max(x)
```

-   Quick way:

```{r}
#| echo: true
#| eval: false
skim(x)
```

## Type 2: `character`

-   For text data

-   Can't use with math calculation, for example, below code will raise error:

```{r}
#| echo: true
#| eval: false
"60" + "8"
```

-   Create text vector

```{r}
#| echo: true
s = c("ann", "betty", "candy")
```

-   Some text manipulation with `stringr` package:

```{r}
#| echo: true
length(s)
str_length(s)
str_to_upper(s)
str_replace(string = s, pattern = "a", replacement = "$")
paste0("student_", s)
```

## Type 3: `factor`

-   Some category only, not continuous numeric variable

-   For example:

```{r}
#| echo: true
f1 = c(0, 1, 1, 0 ,1)
f2 = c("Monday", "Tuesday", "Wednesday", "Tuesday", "Friday")
```

-   Sometimes, we need `factor` to do some special manipulation, e.g., plotting data by groups

## Type 4: `date`

-   Make date:

```{r}
#| echo: true
d = c(20211231, 20221231, 20231231)
typeof(d)
```

-   Convert `numeric` in yyyymmdd format to `date`:

```{r}
#| echo: true
d = ymd(d)
typeof(d)
```

## Missing data: `NA`

-   Sometimes, we do not have data for some cases in the data

-   In data science, people call it missing data

-   In R, it is denoted as `NA` (pronounced as "Non-Applicable")

## Vector vs dataframe

-   Vector: One column of data

    -   E.g., `x = c(5,7,9)` is a vector with three elements, where the first element can be index/choose by `x[1]`

-   Dataframe: Many columns put together in the same table

    -   A dataframe has rows/observations and columns/variables, such as `Data[rows, columns]`

    -   To select first 5 rows rows: `Data[1:5,]`

    -   To select column 2 and 3: `Data[,c(2,3)]`

## Example

```{r}
#| echo: true
mtcars[1:5, 1:4]
mtcars[,1]
mtcars[1,]
```

## Import data

-   Raw data can be in many formats: the most common are

    -   CSV or Excel

    -   Other statistical software: Stata, SAS, SPSS, `parquet`

    -   R native data: `rds`, `RData`

    -   (SQL) Database

-   We will learn first with simple CSV/Excel file. Then, I will introduce when and how to get data from SQL Database

## Import data to be dataframe in R

-   Use interactive import panel

-   Use functions

    -   Depending on data, we have corresponding functions

    -   For example, `read_csv` to read CSV files to R, `read_excel` to read Excel files (we may need to load libraries to use these functions)

## Example

-   Read the file `hsb.csv` to R by two ways

## Save dataframe from R to a file

-   Save output dataframe from R to a file to use later
-   Many possible file types: CSV, `rds`, `parquet`
    -   E.g., `saveRDS(OUTPUT, "path_to_file.rds")`, `readr::write_csv(OUTPUT, "path_to_file.csv")`

## Some tasks with the dataframe

-   Make data smaller

-   Make data larger

-   Don't change the data size

## Make data smaller

| Function                   | When to use                         |
|----------------------------|-------------------------------------|
| `select`                   | Choose a few variables              |
| `filter`                   | Choose few rows                     |
| `group_by` and `summarize` | Summarize some statistics by groups |

## Make data larger

| Function | When to use |
|----|----|
| `mutate` | Create new variables |
| `transmute` | Also create new variables, but keep only these new variables and drop all old variables |

## Don't change the data size

| Function  | When to use             |
|-----------|-------------------------|
| `arrange` | Sort data               |
| `rename`  | To change variable name |

## How to use a function?

-   Google

    -   ChatGPT

-   Use `?` before the function name:

```{r}
#| echo: true
#| eval: false
?mutate
```

-   Practice makes perfect

## Thank you

We are ready to clean a data in the next slide.
